\documentclass[12pt,a4paper]{article}

\usepackage[margin=1in]{geometry}
\usepackage[hidelinks]{hyperref}
\usepackage{verbatim}
\usepackage{longtable}
\usepackage{listings}

\lstset{
  basicstyle=\ttfamily\scriptsize,
  breaklines=true,
  breakatwhitespace=false,
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  xleftmargin=0pt,
  xrightmargin=0pt
}

\begin{document}

\begin{titlepage}
    \centering
    \vspace*{1.5cm}
    
    {\Huge \textbf{University of Engineering and Technology}}\\[2cm]
    
    {\LARGE \textbf{Compiler Construction Lab}}\\[0.5cm]
    {\Large \textbf{Mini Compiler Project Report}}\\[2.5cm]
    
    {\Large \textbf{Submitted By:}}\\[0.5cm]
    {\large Member 1: Munees Tariq (2023-CS-32)}\\[0.2cm]
    {\large Member 2: Muhammad Mahad Sheraz (2023-CS-36)}\\[1.5cm]
    
    {\large \textbf{Section:}} A\\[2cm]
    
    {\Large \textbf{Supervisor:}}\\[0.5cm]
    {\large Mam Aliya Farooq}\\[2cm]
    
    \vfill
    
    {\large \textbf{Spring 2026}}
\end{titlepage}

\tableofcontents
\newpage

\section{Introduction}
This report details the design and implementation of the mini compiler for the CS-471L Compiler Construction Lab final project. The compiler is built for a subset of the MicroJava programming language. It integrates all the lab modules developed during the semester, along with additional components such as an LR Parser, Symbol Table Manager, and Error Handling.

\section{Grammar Specification}
The language subset used is MicroJava. The grammar supports class declarations, variable declarations, methods, statements (if, while, return, print), expressions, and assignments. 
The original grammar, as well as the left-factored and non-left-recursive version used for top-down parsing, are specified in the accompanying documentation.

\subsection{Original and LL(1) Grammar}
The full grammar in BNF form can be found in \texttt{docs/grammar.md}.
\lstinputlisting{docs/grammar.md}

\section{Module Descriptions}

\subsection{Lexical Analyzer (Lab 3)}
The lexical analyzer (\texttt{src/lexer.c}) is responsible for tokenizing the input source program. It reads characters using a double buffering technique and groups them into logical tokens such as keywords, identifiers, numbers, operators, and punctuation. 
It ignores whitespace and comments and keeps track of the line and column number for each token to assist in error reporting. 

\subsection{Recursive Descent Parser (Lab 4)}
The recursive descent parser (\texttt{src/recursive\_descent\_parser.c}) is a top-down parser implementing a set of mutually recursive functions, one for each non-terminal in the grammar. It uses the left-factored and non-left-recursive version of the grammar to avoid infinite loops and ambiguity.

\subsection{Non-Recursive Predictive Parser (Lab 5)}
The predictive parser (\texttt{src/predictive\_parser.c}) is an LL(1) top-down parser that avoids recursion by using an explicit stack and a parsing table. 
The FIRST and FOLLOW sets used to construct the LL(1) parsing table can be found in \texttt{docs/first\_follow.md}, and the table itself is in \texttt{docs/ll1\_table.md}.

\subsection{LR Parser (New Module)}
The LR parser (\texttt{src/lr\_parser.c}) is a bottom-up SLR(1) parser. It uses a stack of states and grammar symbols alongside an action and goto table.
It parses the original grammar (including left-recursive rules) by performing shift and reduce operations. 

\subsection{Symbol Table Manager (Lab 7, Extended)}
The symbol table (\texttt{src/symbol\_table.c}) uses a hash-based structure for efficient lookups. It handles nested scopes using a linked scope chain. For each identifier, it stores attributes such as name, kind (e.g., variable, class, function), type, scope level, and source line number. It is integrated with the parser to populate declarations and resolve usages.

\subsection{Error Handler (Bonus Task)}
The error handler (\texttt{src/error\_handler.c}) centrally manages lexical, syntactic, and semantic errors. It implements panic-mode recovery, preventing the parser from crashing on the first error and allowing it to report multiple issues in a single pass. Errors include detailed line and column numbers.

\section{Construction of the LR Parsing Table}
The SLR(1) parsing technique was chosen for the LR parser.
The steps to construct the SLR(1) parsing table are as follows:
\begin{enumerate}
    \item \textbf{Augment the Grammar:} A new start symbol is added to the grammar to explicitly define when the parser should accept the input.
    \item \textbf{LR(0) Item Sets:} We construct the canonical collection of LR(0) items, determining the states of the parser.
    \item \textbf{Goto Graph:} Transitions between the item sets on terminals and non-terminals are defined.
    \item \textbf{Action Table:}
    \begin{itemize}
        \item Shift actions are added for terminal transitions.
        \item Reduce actions are added for completed items, but only for lookahead symbols present in the FOLLOW set of the production's left-hand side non-terminal.
        \item An accept action is added for the augmented start production.
    \end{itemize}
    \item \textbf{Goto Table:} Non-terminal transitions are added to the Goto table.
\end{enumerate}

The full LR Action and Goto tables are available in \texttt{docs/lr\_table.md}.

\section{Module Integration}
All modules are integrated into a single unified build system. The file \texttt{src/compiler.c} acts as the main entry point (dispatcher). It processes command-line arguments to determine which module to run (lexer, recursive, predictive, lr, or all). 
The lexer provides a token stream that any of the three parsers can consume. The parsers communicate with the Symbol Table Manager for semantic checks and the Error Handler to report issues gracefully.
The compiler can be built using the provided \texttt{Makefile} or \texttt{build.bat} script.

\section{Sample Input and Output}

\subsection{Sample Input}
A valid MicroJava program, \texttt{hello.mj}:
\lstinputlisting{test/samples/hello.mj}

\subsection{Invalid Syntax Input}
\lstinputlisting{test/samples/invalid_missing_semicolon.mj}

\subsection{Lexer Output}
\begin{verbatim}
1:1     KEYWORD 'class' 'class'
1:7     IDENTIFIER      'Hello' 'Hello'
1:13    PUNCTUATION     '{'     '{'
2:3     KEYWORD 'void'  'void'
2:8     IDENTIFIER      'main'  'main'
2:12    PUNCTUATION     '('     '('
2:13    PUNCTUATION     ')'     ')'
2:15    PUNCTUATION     '{'     '{'
3:5     KEYWORD 'print' 'print'
3:10    PUNCTUATION     '('     '('
3:11    NUMBER  '1'     '1'
3:13    OPERATOR        '+'     '+'
3:15    NUMBER  '2'     '2'
3:17    OPERATOR        '*'     '*'
3:19    NUMBER  '3'     '3'
3:20    PUNCTUATION     ')'     ')'
3:21    PUNCTUATION     ';'     ';'
4:5     KEYWORD 'return'        'return'
4:11    PUNCTUATION     ';'     ';'
5:3     PUNCTUATION     '}'     '}'
6:1     PUNCTUATION     '}'     '}'
7:1     EOF     ''      NULL
\end{verbatim}

\subsection{LR Parser Trace}
For at least one sample input, we show a trace with the stack contents, the remaining input, and the action taken at each step. See \texttt{docs/hello\_lr\_trace\_utf8.txt} for a detailed trace.
\lstinputlisting{docs/hello_lr_trace_utf8.txt}

\subsection{Symbol Table Output}
When semantic analysis is enabled on \texttt{hello.mj}:
\begin{verbatim}
Symbol table dump
Name                 Kind       Type       Scope  Line
main                 function   void       0      2
Hello                class      class      0      1
\end{verbatim}

\section{Limitations}
\begin{itemize}
    \item The compiler is a front-end only; it does not generate intermediate representation or target machine code.
    \item Semantic checking ensures proper scope rules and declarations but does not implement comprehensive type checking for complex expressions.
    \item Error recovery uses basic panic-mode techniques (skipping tokens until a synchronization point). More advanced phrase-level recovery is limited.
\end{itemize}

\section{References}
\begin{enumerate}
    \item Aho, A. V., Lam, M. S., Sethi, R., \& Ullman, J. D. (2006). \textit{Compilers: Principles, Techniques, and Tools} (2nd ed.). Pearson.
    \item Course Materials and Lab Manuals provided by Instructor Aliya Farooq.
\end{enumerate}

\end{document}
